---
title: "A UN Report on the Relationship between GDP and Life Expectancy"
author: "Patrick D. Schloss"
date: "2023-10-13"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(here)
library(knitr)

options(scipen = 1000000)
```

## Abstract

This report was prepared to the attention of the UN. It analyzes the relationship between a country's GDP, its life expectancy, and CO2 emissions. Our goal is to determine to what degree a country's economic strength or weakness is related to its public health status and impact on climate pollution. We hypothesize that both life expectancy and CO2 emissions increase with the country's GDP.

```{r data, include = FALSE}
gapminder_1997 <- read_csv(here("data/gapminder_1997.csv"))
```

## Plotting GDP by Life Expectancy

```{r gdp_lifeexp_1997, echo = FALSE}
ggplot(data = gapminder_1997, 
       aes(x = gdpPercap, y = lifeExp,
           color = continent, size = pop/1000000)) +
  geom_point() +
  labs(x = "GDP per Capita",
       y= "Life Expectancy",
       title = "Do people in wealthy countries live longer?",
       size = "Population (in millions)")
```

```{r summary_stats, include = FALSE}
n_countries <- gapminder_1997 %>%
  select(country) %>%
  n_distinct()
  
minimum_gdp <- gapminder_1997 %>%
  summarize(min(gdpPercap)) %>%
  round(digits = 1)

maximum_gdp <- gapminder_1997 %>%
  summarize(max(gdpPercap)) %>%
  round(digits = 1)

# format - will format things nicely
```

## Summary statistics

The above plot shows the relationship between GDP per capita and life expectancy for a total of `r n_countries` countries. For this set of countries, economic wealth ranged from a minimum of \$`r minimum_gdp` to a maximum of \$`r maximum_gdp` per capita. $n=2+2$


| Summary of data  |
|:----------|-----------:|
|Number of countries | `r n_countries` |
|Minimum GDP per capita | `r minimum_gdp` |
|Maximum GDP per capita | `r maximum_gdp` |


```{r table, echo = FALSE}
gapminder_1997 %>%
  filter(continent == "Oceania") %>%
  kable()
```

## Exercises

#### Exercise 1:
Try googling how to create sections by using headers and sub-headers using Markdown. What do you find?

### Exercise 2:
Go ahead and do some online searches on how to do the following:

* Create a bullet point list with three items
* As the first item, write the name of your currently favorite programming language in bold
* As the second item, write the name of a function you have so far found most useful in italics
* As the third item, write one thing you want to learn next on your programming journey in bold and italics
* Turn your bullet point list into a numbered list
* Create a fourth list item and find an online guide and/or cheat sheet for basic Markdown syntax, write its name down here and hyperlink its URL

My bulleted list

* My favorite programming language is **R**
* The most useful function has been *group_by*
* The next thing I want to learn is ***RNAseq analysis***

My numbered list

1. My favorite programming language is **R**
1. The next thing I want to learn is ***RNAseq analysis***
1. The most useful function has been *group_by*
1. This is my [favorite markdown cheatsheet](https://rstudio.github.io/cheatsheets/html/rmarkdown.html?_gl=1*l8arwq*_ga*ODA0MTA1NDY2LjE2OTcyMjQwNTk.*_ga_2C0WZ1JHG0*MTY5NzIyNDA1OS4xLjAuMTY5NzIyNDA1OS4wLjAuMA..)
